### 2018-06-16 paper/2-background.tex
Fix paper build
Finish S5.2

\Make\footnote{There are numerous implementations of \Make and none comes with a
formal specification. In this paper we therefore use a simple and sensible
approximation to a real \Make that you might find on your machine.} was developed
more than 40 years ago to automatically build software libraries and executable
programs from source code. It uses \emph{makefiles} to describe tasks (often
referred to as \emph{build rules}) and their dependencies in a simple textual form.
For example\footnote{Do not worry about the dependency on \cmd{stdio.h}, and its transitive (perhaps dynamic) dependencies.  We will deal with these in Section \simon{xxx}.}:

@@

To support dynamic dependencies,
\Excel's build algorithm~\cite{excel_recalc} is significantly different from
\Make. \Excel uses the \emph{calculation chain}
produced by the previous build as an approximation to the correct topological
order. During recalculation, \Excel processes cells in this order, but can
\emph{defer recalculation of a cell} by moving it down the chain if a newly
discovered dependency has not yet been rebuilt. We refer to this
algorithm as \emph{reordering}, and will discuss it in more detail
in~\S\ref{sec-implementation-excel}.

Dynamic dependencies complicate minimality.  In the above example,
\cmd{B1} should only be recomputed if \cmd{A1} or \cmd{C1} change, but
not if (say) \cmd{A2} changes; but these facts are not statically apparent.
In pracice Excel implements a conservative approximation to
minimality: roughly, it recomputes a formula if (a) the formula
text mentions a changed cell, or (b) the formula uses a function like \cmd{INDIRECT}
whose dependencies are not statically visible.

>>>>

\Make\footnote{There are numerous implementations of \Make and none comes with a
formal specification. In this paper we therefore use a simple and sensible
approximation to a real \Make that you might find on your machine.} was developed
more than 40 years ago to automatically build software libraries and executable
programs from source code. It uses \emph{makefiles} to describe tasks (often
referred to as \emph{build rules}) and their dependencies in a simple textual form.
For example\footnote{Do not worry about the dependency on \cmd{stdio.h}, and
its transitive (perhaps dynamic) dependencies.
We will deal with these in Section~\S\ref{sec-task-monad}.}:

@@

To support dynamic dependencies,
\Excel's build algorithm~\cite{excel_recalc} is significantly different from
\Make. \Excel uses the \emph{calculation chain}
produced by the previous build as an approximation to the correct topological
order. During recalculation, \Excel processes cells in this order, but can
\emph{defer recalculation of a cell} by moving it down the chain if a newly
discovered dependency has not yet been rebuilt. We refer to this
algorithm as \emph{restarting}, and will discuss it in more detail
in~\S\ref{sec-implementation-excel}.

Dynamic dependencies complicate minimality.  In the above example, \cmd{B1}
should only be recomputed if \cmd{A1} or \cmd{C1} change, but not if (say)
\cmd{A2} changes; but these facts are not statically apparent. In practice
\Excel implements a conservative approximation to minimality: roughly, it
recomputes a formula if (i) the formula text mentions a changed cell, or (ii)
the formula uses a function like \cmd{INDIRECT} whose dependencies are not
statically visible.

### 2018-06-16 paper/5-implementations.tex
Finish S5.2

\begin{figure}
\begin{minted}[fontsize=\small]{haskell}
-- Make build system
make :: Ord k => Build Applicative (MakeInfo k) k v
make = topological modTimeRebuilder
\end{minted}
\vspace{0mm}
\begin{minted}[fontsize=\small]{haskell}
-- Build information: current time and file modification times
type Time       = Integer
type MakeInfo k = (Time, Map k Time)
\end{minted}
\vspace{0mm}
\begin{minted}[fontsize=\small]{haskell}
-- A task rebuilder based on file modification times
modTimeRebuilder :: Ord k => Rebuilder Applicative (MakeInfo k) k v
modTimeRebuilder key value task = Task $ \fetch -> do
    (now, modTime) <- get
    let dirty = case Map.lookup key modTime of
            Nothing -> True
            time -> any (\d -> Map.lookup d modTime > time) (dependencies task)
    if not dirty then return value else do
        put (now + 1, Map.insert key now modTime)
        run task fetch
\end{minted}
\vspace{0mm}
\begin{minted}[fontsize=\small]{haskell}
-- A topological task scheduler
topological :: Ord k => Rebuilder Applicative i k v -> Build Applicative i k v
topological rebuilder tasks target = execState $ forM_ order $ \key -> case tasks key of
    Nothing   -> return ()
    Just task -> do
        value <- gets (getValue key)
        let newTask :: Task (MonadState i) k v
            newTask = rebuilder key value task
            fetch :: k -> StateT i (State (Store i k v)) v
            fetch = lift . gets . getValue
        info <- gets getInfo
        (newValue, newInfo) <- runStateT (run newTask fetch) info
        modify $ putInfo newInfo . putValue key newValue
  where
    deps k = case tasks k of { Nothing -> [@@]; Just task -> dependencies task }
    order  = case topSort (reachable deps target) of
        Nothing -> error "Cannot build tasks with cyclic dependencies"
        Just xs -> xs
\end{minted}
\vspace{0mm}
\begin{minted}[fontsize=\small]{haskell}
-- Standard graph algorithms (implementation omitted)
reachable :: Ord k => (k -> [k]) -> k -> Graph k
topSort   :: Ord k => Graph k -> Maybe [k]
\end{minted}
\vspace{-2mm}
\caption{An implementation of \Make using our framework.}\label{fig-make-implementation}
\vspace{-4mm}
\end{figure}

@@

\Make's scheduler processes keys in a linear \hs{order} based on a
\hs{topological} sort of the statically known dependency graph
(see~\S\ref{sec-parallelism} for parallel \Make). Our definition in
Fig.~\ref{fig-make-implementation} is polymorphic with respect to the type of
build information \hs{i} and is therefore compatible with any applicative
\hs{rebuilder}. The scheduler calls the supplied \hs{rebuilder} on every
\hs{key} in the \hs{order}, and runs the obtained \hs{newTask} to compute the
\hs{newValue} and updated build information \hs{newInfo}, which are subsequently
written to the store. The computation runs in the
\hs{State}~\hs{(@@Store}~\hs{i}~\hs{k}~\hs{v)} monad, similar to the \hs{busy}
build system (\S\ref{sec-general-build}), but we need to use an additional
\hs{StateT}~\hs{i} wrapper to run the \hs{newTask} since it only has access to
the \hs{i} part of the \hs{Store}~\hs{i}~\hs{k}~\hs{v}. The \hs{newTask} can
find values of the \hs{key}'s dependencies via the \hs{fetch} callback that
reads the store by executing \hs{getValue} in the inner state monad.

@@

Fig.~\ref{fig-excel-implementation} shows an implementation of \Excel in our
framework as a combination of the \hs{restarting} scheduler and an
\hs{approximateRebuilder}.

\Excel's persistently stored information is a triple: (i) the set of dirty keys
\hs{Set}~\hs{k}, (ii) a map storing known upper bounds of dependencies
\hs{Map}~\hs{k}~\hs{[@@k]} that \Excel uses to handle dynamic dependencies, and
(iii) the calculation chain \hs{[@@k]} recorded in the previous build
(\S\ref{sec-background-excel}).

\begin{figure}
\begin{minted}[fontsize=\small]{haskell}
-- Excel build system
excel :: Ord k => Build Monad (ExcelInfo k) k v
excel = restarting approximateRebuilder
\end{minted}
\vspace{1mm}
\begin{minted}[fontsize=\small]{haskell}
-- Approximate task dependencies: a map of known upper bounds of dependencies
type ApproximateDependencies k = Map k [k]
\end{minted}
\vspace{1mm}
\begin{minted}[fontsize=\small]{haskell}
-- A task rebuilder based on the set of dirty keys and approximate dependencies
approximateRebuilder :: Ord k => Rebuilder Monad (Set k, ApproximateDependencies k) k v
approximateRebuilder key value task = Task $ \fetch -> do
    (isDirty, deps) <- get
    let dirty = key `Set.member` isDirty || case Map.lookup key deps of
                                                Nothing -> True
                                                Just ks -> any (`Set.member` isDirty) ks
    if not dirty then return value else do
        put (Set.insert key isDirty, deps)
        run task fetch
\end{minted}
\vspace{1mm}
\begin{minted}[fontsize=\small]{haskell}
-- Build information: dirty keys, approximate dependencies, and calculation chain
type Chain k = [k]
type ExcelInfo k = ((Set k, ApproximateDependencies k), Chain k)
\end{minted}
\vspace{1mm}
\begin{minted}[fontsize=\small]{haskell}
-- A restarting task scheduler
restarting :: Ord k => Rebuilder Monad i k v -> Build Monad (i, Chain k) k v
restarting rebuilder tasks key = execState $ do
    chain    <- gets (snd . getInfo)
    newChain <- go Set.empty $ chain ++ [key | key `notElem` chain]
    modify . mapInfo $ \(i, _) -> (i, newChain)
  where
    go :: Set k -> Chain k -> State (Store (i, Chain k) k v) (Chain k)
    go _    []     = return []
    go done (k:ks) = case tasks k of
        Nothing -> (k :) <$> go (Set.insert k done) ks
        Just task -> do
            store <- get
            let value = getValue k store
                newTask :: Task (MonadState i) k (Either k v)
                newTask = try $ rebuilder k value task -- `try' is defined below
                fetch :: k -> State i (Either k v)
                fetch k | k `Set.member` done = return $ Right (getValue k store)
                        | otherwise           = return $ Left k
            case runState (run newTask fetch) (fst $ getInfo store) of
                (Left dep, _) -> go done $ [ dep | dep `notElem` ks ] ++ ks ++ [k]
                (Right newValue, newInfo) -> do
                    modify $ putInfo (newInfo, []) . updateValue k value newValue
                    (k :) <$> go (Set.insert k done) ks
\end{minted}
\vspace{1mm}
\begin{minted}[fontsize=\small]{haskell}
-- Convert a total task into a task accepting a partial fetch callback
try :: Task (MonadState i) k v -> Task (MonadState i) k (Either e v)
try task = Task $ \fetch -> runExceptT $ run task (ExceptT . fetch)
\end{minted}
\vspace{-2mm}
\caption{An implementation of \Excel using our framework.}\label{fig-excel-implementation}
\vspace{-2mm}
\end{figure}

The helper function \hs{reordering}, whose implementation we omit since it is
technical and not particularly enlightening, calls the function \hs{process} to
\emph{try to build} a \hs{key} by executing the action \hs{act}, in the order
determined by the calculation chain. To decide whether to \hs{rebuild} the
\hs{key}, \hs{process} checks if the \hs{key} itself is marked dirty or the
approximation of its dependencies contains a dirty key. Notice that if the
dependencies of the \hs{key} are \hs{Unknown}, e.g. when it uses the
\cmd{INDIRECT} function, the \hs{key} is always rebuilt. If a \hs{rebuild} is
not needed, \hs{process} returns \hs{Nothing} to indicate that. Otherwise, it
executes \hs{act} leading to one of two possible \hs{result}s:

\begin{itemize}
    \item \hs{MissingDependency}~\hs{k} indicates that the execution of \hs{act} has
    failed, because one of its dependencies \hs{k} was out-of-date, i.e. the
    calculation chain from the previous build was incorrect and therefore needs
    to be reordered, deferring the \hs{key} to be rebuilt later.
    \item \hs{Result}~\hs{v}~\hs{dynamicDependencies} indicates that the
    execution has succeeded producing the value~\hs{v} and the list of the
    \hs{key}'s dynamic dependencies. We store the value, and mark it dirty to
    trigger the rebuilding of keys that depend on it.
\end{itemize}

In both of the above cases, we notify the parent \hs{reordering} function of
the outcome by returning \hs{Just}~\hs{result}. The astute reader may notice
that \hs{process} ignores \hs{dynamicDependencies} available in the \hs{result}.
While not required for \Excel, we have implemented build systems using
\hs{reordering} which use verifying and constructive traces, effectively turning
\Excel into a cloud build system and ensuring \hs{reordering} is not overly
fitted to \Excel alone.
%  (additional implementations will be submitted to the
% Artefact Evaluation).

In reality \Excel works slightly differently (as far as we are able to
ascertain)~--~on a change it propagates the dirty bits forward using the
\hs{dynamicDependencies}, then only checks if the current \hs{key} is dirty.
While both methods are equivalent, merely changing the interleaving, our
approach allows us to model more of the behaviour of \Excel.

>>>>

\begin{figure}
\begin{minted}[fontsize=\small]{haskell}
-- Make build system
make :: Ord k => Build Applicative (MakeInfo k) k v
make = topological modTimeRebuilder
\end{minted}
\vspace{0mm}
\begin{minted}[fontsize=\small]{haskell}
-- Build information: current time and file modification times
type Time       = Integer
type MakeInfo k = (Time, Map k Time)
\end{minted}
\vspace{0mm}
\begin{minted}[fontsize=\small]{haskell}
-- A task rebuilder based on file modification times
modTimeRebuilder :: Ord k => Rebuilder Applicative (MakeInfo k) k v
modTimeRebuilder key value task = Task $ \fetch -> do
    (now, modTime) <- get
    let dirty = case Map.lookup key modTime of
            Nothing -> True
            time -> any (\d -> Map.lookup d modTime > time) (dependencies task)
    if not dirty then return value else do
        put (now + 1, Map.insert key now modTime)
        run task fetch
\end{minted}
\vspace{0mm}
\begin{minted}[fontsize=\small]{haskell}
-- A topological task scheduler
topological :: Ord k => Rebuilder Applicative i k v -> Build Applicative i k v
topological rebuilder tasks target = execState $ forM_ order $ \key -> case tasks key of
    Nothing   -> return ()
    Just task -> do
        store <- get
        let newTask :: Task (MonadState i) k v
            newTask = rebuilder key (getValue key store) task
            fetch :: k -> State i v
            fetch k = pure (getValue k store)
            (newValue, newInfo) = runState (run newTask fetch) (getInfo store)
        modify $ putInfo newInfo . putValue key newValue
  where
    deps k = case tasks k of { Nothing -> [@@]; Just task -> dependencies task }
    order  = case topSort (reachable deps target) of
        Nothing -> error "Cannot build tasks with cyclic dependencies"
        Just xs -> xs
\end{minted}
\vspace{0mm}
\begin{minted}[fontsize=\small]{haskell}
-- Standard graph algorithms (implementation omitted)
reachable :: Ord k => (k -> [k]) -> k -> Graph k
topSort   :: Ord k => Graph k -> Maybe [k]
\end{minted}
\vspace{-2mm}
\caption{An implementation of \Make using our framework.}\label{fig-make-implementation}
\vspace{-4mm}
\end{figure}

@@

\Make's scheduler processes keys in a linear \hs{order} based on a
\hs{topological} sort of the statically known dependency graph
(see~\S\ref{sec-parallelism} for parallel \Make). Our definition in
Fig.~\ref{fig-make-implementation} is polymorphic with respect to the type of
build information \hs{i} and is therefore compatible with any applicative
\hs{rebuilder}. The scheduler calls the supplied \hs{rebuilder} on every
\hs{key} in the \hs{order}, and runs the obtained \hs{newTask} to compute the
\hs{newValue} and updated build information \hs{newInfo}, which are subsequently
written to the store. The computation runs in the
\hs{State}~\hs{(@@Store}~\hs{i}~\hs{k}~\hs{v)} monad, similar to the \hs{busy}
build system (\S\ref{sec-general-build}), however since \hs{newTask} has access
only to the \hs{i} part of the \hs{Store}~\hs{i}~\hs{k}~\hs{v}, we need to use
\hs{runState} to execute it. The \hs{newTask} finds values of the \hs{key}'s
dependencies via the \hs{fetch} callback that directly reads the \hs{store}.

@@

Fig.~\ref{fig-excel-implementation} shows an implementation of \Excel in our
framework as a composition of the \hs{restarting} scheduler and an
\hs{approximateRebuilder}.

\Excel's persistently stored build information is a triple:

\begin{itemize}
    \item A set of dirty keys \hs{Set}~\hs{k}, i.e. all cells whose contents
          has changed since the previous build, including cells that contain
          input values as well as formulas.
    \item A map \hs{Map}~\hs{k}~\hs{[@@k]} that stores known upper bounds of
          dependencies. If there is an entry for a key, it is a conservative
          approximation of its dependencies (e.g. the dependencies of the
          formula \cmd{IF(C1=1,B2,A2)} can be approximated by the list
          \hs{[}\cmd{C1}\hs{,}\cmd{B2}\hs{,}\cmd{A2}\hs{]}). Otherwise, there is
          no upper bound (e.g. the key uses an \cmd{INDIRECT} reference) and
          the key is assumed to be dirty.
    \item A calculation \hs{Chain} recorded in the previous build
          (\S\ref{sec-background-excel}), represented by a list of keys \hs{[@@k]}.
\end{itemize}

The \hs{approximateRebuilder} uses the first two elements of the above triple
to decide whether a given \hs{key} is \hs{dirty} and hence needs to be rebuilt.
The logic is similar to that of \Make's \hs{modTimeRebuilder}, but instead of
relying on file modification times, \Excel stores the information about dirty
keys more directly. Furthermore, \Excel's dependency approximation makes it
possible to build tasks with dynamic dependencies: a \hs{key} is declared
\hs{dirty} if any of its approximate dependencies is in the \hs{dirtyKeys} set,
or if there is no known upper bound on dependencies, i.e. if the lookup in the
map \hs{deps}~\hs{::}~\hs{Map}~\hs{k}~\hs{[@@k]} fails. Note that when a
\hs{key} is rebuilt it is added to the \hs{dirtyKeys} set to trigger further
rebuilds.

\Excel's \hs{restarting} scheduler processes keys in the order specified by the
calculation \hs{chain}, constructing a \hs{newChain} for the next build and
maintaining a set of keys \hs{done} that have already been processed. For each
non-input \hs{key}, the scheduler tries to rebuild it using a partial \hs{fetch}
callback that returns \hs{Either}~\hs{k}~\hs{v} instead of \hs{v}. The callback
is defined to fail with \hs{Left}~\hs{dep} when asked for the value of a
dependency \hs{dep} that has not yet been processed (and hence may potentially
be dirty); otherwise it returns the current value of the dependency in the
\hs{store}.

After the \hs{newTask} is executed with \hs{runState} there are two cases to
consider:

\begin{itemize}
    \item The \hs{newTask} has failed, because one of its dependencies \hs{dep}
    has not yet been processed. This indicates that the calculation \hs{chain}
    from the previous build is incorrect and needs to be adjusted by making sure
    it contains \hs{dep} and moving the \hs{key} to the end, so that its build
    can be restarted after the \hs{dep} is ready.
    \item The \hs{newTask} succeeded. The resulting \hs{newValue} and
    \hs{newInto} are written to the store, and \Excel continues to build the
    \hs{chain} after marking the \hs{key} as \hs{done}.
\end{itemize}

Note that the task returned by the \hs{rebuilder} expects a total callback
function and cannot be directly executed with the partial callback \hs{fetch}.
We fix the mismatch with the function \hs{try} that relies on the standard
monad transformer \hs{ExceptT} from the \cmd{transformers} library.

\begin{figure}
\begin{minted}[fontsize=\small]{haskell}
-- Excel build system
excel :: Ord k => Build Monad (ExcelInfo k) k v
excel = restarting approximateRebuilder
\end{minted}
\vspace{0mm}
\begin{minted}[fontsize=\small]{haskell}
-- Build information: dirty keys, approximate dependencies, and calculation chain
type Chain k = [k]
type ExcelInfo k = ((Set k, Map k [k]), Chain k)
\end{minted}
\vspace{0mm}
\begin{minted}[fontsize=\small]{haskell}
-- A task rebuilder based on the set of dirty keys and approximate dependencies
approximateRebuilder :: Ord k => Rebuilder Monad (Set k, Map k [k]) k v
approximateRebuilder key value task = Task $ \fetch -> do
    (dirtyKeys, deps) <- get
    let dirty = key `Set.member` dirtyKeys ||
                case Map.lookup key deps of Nothing -> True
                                            Just ks -> any (`Set.member` dirtyKeys) ks
    if not dirty then return value else do
        put (Set.insert key dirtyKeys, deps)
        run task fetch
\end{minted}
\vspace{0mm}
\begin{minted}[fontsize=\small]{haskell}
-- A restarting task scheduler
restarting :: Ord k => Rebuilder Monad i k v -> Build Monad (i, Chain k) k v
restarting rebuilder tasks target = execState $ do
    chain    <- gets (snd . getInfo)
    newChain <- go Set.empty $ chain ++ [target | target `notElem` chain]
    modify . mapInfo $ \(i, _) -> (i, newChain)
  where
    go :: Set k -> Chain k -> State (Store (i, Chain k) k v) (Chain k)
    go _    [@@]       = return [@@]
    go done (key:ks) = case tasks key of
        Nothing -> (key :) <$> go (Set.insert key done) ks
        Just task -> do
            store <- get
            let newTask :: Task (MonadState i) k (Either k v)
                newTask = try $ rebuilder key (getValue key store) task
                fetch :: k -> State i (Either k v)
                fetch k | k `Set.member` done = return $ Right (getValue k store)
                        | otherwise           = return $ Left k
            case runState (run newTask fetch) (fst $ getInfo store) of
                (Left dep, _) -> go done $ [ dep | dep `notElem` ks ] ++ ks ++ [key]
                (Right newValue, newInfo) -> do
                    modify $ putInfo (newInfo, [@@]) . putValue key newValue
                    (key :) <$> go (Set.insert key done) ks
\end{minted}
\vspace{0mm}
\begin{minted}[fontsize=\small]{haskell}
-- Convert a total task into a task accepting a partial fetch callback
try :: Task (MonadState i) k v -> Task (MonadState i) k (Either e v)
try task = Task $ \fetch -> runExceptT $ run task (ExceptT . fetch)
\end{minted}
\vspace{-2mm}
\caption{An implementation of \Excel using our framework.}\label{fig-excel-implementation}
\vspace{-4mm}
\end{figure}
